---
title: "Red Wine Quality Exploration"
author: "Eric"
date: "October 24, 2017"
output: html_document
---

Red Wine Quality Exploration by Eric
========================================================

"Tasty like a fine wine" - ever hear this expression before? Ever wonder what
makes wine, a fermented grape juice by nature, taste better with age? Does all
wine taste better with age? While the age of wine is interesting in measuring 
the quality of it, this data exploration project will dive into the chemical
properties of red wine and explore what makes a good red wine better than others.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

#install.packages('ggplot2', dependencies = T)
#install.packages('gridExtra')
#install.packages('GGally')
#install.packages('scales')
#install.packages('memisc')
#install.packages('lattice')
#install.packages('MASS')
#install.packages('car')
#install.packages('reshape')
#install.packages('dplyr')
#install.packages('RColorBrewer')
library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine_df <- read.csv('wineQualityReds.csv')

#unique(wine_df$quality)

wine_df$quality_rank <- factor(wine_df$quality, ordered = T)

# assign ratings to the quality rank
wine_df$quality_rank <- ifelse(wine_df$quality < 5, 'poor', 
                               ifelse(wine_df$quality < 7, 'average', 
                                      'excellent'))

wine_df$quality_rank <- ordered(wine_df$quality_rank,
                                levels = c('poor', 'average', 'excellent'))
```

# Univariate Plots Section

## About the Data

This report explores over 1500 observations containing information about 12 
different attributes of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_ExploreData}
# explore the data set feature set
names(wine_df)
summary(wine_df)
str(wine_df)
```

## Wine Quality Distribution

The minimum value of the quality of wine is a 3 and the highest scored 
observation received an 8 for wine quality. The median quality score equals 6 
and the bulk of the quality scores fall within the [5,6] range.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Quality}
# explore quality - [3,8]
# try histogram
ggplot(aes(x = quality), data = wine_df) +
  geom_histogram(binwidth = .5, 
                 color = I('black'), 
                 fill = I('#099DD9'))

# histogram for quality_rank
ggplot(aes(x = quality_rank), data = wine_df) +
  geom_bar(width = .5, 
                 color = I('black'), 
                 fill = I('#099DD9'))
```

One thing that stood out in these plots of the wine quality is that the majority 
of the observations fall into the average wine quality bucket. This suggests the 
data set might not be the most trusted data as the majority of wine tasted could 
have been in the medium price range or did the data get collected at a specific 
location (ie a gas station or convenience store)? A more concrete concern with 
this data is how the poor and excellent wine qualities have low observation 
counts. These factors will have to go through some sort of normalization in 
order to smooth out the distributions.

## Distributions For the Remaining Attributes

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
p1 <- ggplot(aes(x = fixed.acidity), data = wine_df) +
  geom_histogram()

p2 <- ggplot(aes(x = volatile.acidity), data = wine_df) +
  geom_histogram()

p3 <- ggplot(aes(x = citric.acid), data = wine_df) +
  geom_histogram()

p4 <- ggplot(aes(x = residual.sugar), data = wine_df) +
  geom_histogram()

p5 <- ggplot(aes(x = chlorides), data = wine_df) +
  geom_histogram()

p6 <- ggplot(aes(x = free.sulfur.dioxide), data = wine_df) +
  geom_histogram()

p7 <- ggplot(aes(x = total.sulfur.dioxide), data = wine_df) +
  geom_histogram()

p8 <- ggplot(aes(x = density), data = wine_df) +
  geom_histogram()

p9 <- ggplot(aes(x = pH), data = wine_df) +
  geom_histogram()

p10 <- ggplot(aes(x = sulphates), data = wine_df) +
  geom_histogram()

p11 <- ggplot(aes(x = alcohol), data = wine_df) +
  geom_histogram()

p12 <- ggplot(aes(x = quality), data = wine_df) +
  geom_histogram()

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12, ncol = 3)
```


# Univariate Analysis

The univariate plots shed light to the behavior of the various chemical 
properties of the wine data.

## Structure of the Dataset

- The following properties are normally distributed: Volatile acidity, Density, 
pH
- The following properties are positively skewed: Fixed acidity, Citric acid, 
Free sulfur dioxide, Total sulfur dioxide, Sulphates, Alcohol
- The following properties are long-tailed skewed: Residual sugar, Chlorides

## Main Feature of Interest

Main feature of interest in the data is quality of the wine. Much of the 
bivariate analysis should be centered around this feature as the independent 
variable. This exploration will hopefully lead to an understanding of which 
factors contribute / how they contribute to the quality of wine.

## Unusual Distributions Investigated

For the features following a skewed or long-tailed distribution, I performed 
arithmetic operations on the data in order to transform them closer to a normal 
distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE, TransformingUnusualDistributions}
# use sqrt for the transformation of positively skewed features
wine_df$fixed.acidity_sqrt <- sqrt(wine_df$fixed.acidity)
p1 <- ggplot(aes(x = fixed.acidity_sqrt), data = wine_df) +
  geom_histogram()

wine_df$citric.acid_sqrt <- sqrt(wine_df$citric.acid)
p3 <- ggplot(aes(x = citric.acid_sqrt), data = wine_df) +
  geom_histogram()

wine_df$free.sulfur.dioxide_sqrt <- sqrt(wine_df$free.sulfur.dioxide)
p6 <- ggplot(aes(x = free.sulfur.dioxide_sqrt), data = wine_df) +
  geom_histogram()

wine_df$total.sulfur.dioxide_sqrt <- sqrt(wine_df$total.sulfur.dioxide)
p7 <- ggplot(aes(x = total.sulfur.dioxide_sqrt), data = wine_df) +
  geom_histogram()

wine_df$sulphates_sqrt <- sqrt(wine_df$sulphates)
p10 <- ggplot(aes(x = sulphates_sqrt), data = wine_df) +
  geom_histogram()

wine_df$alcohol_sqrt <- sqrt(wine_df$alcohol)
p11 <- ggplot(aes(x = alcohol_sqrt), data = wine_df) +
  geom_histogram()

# use log10 for transformation of long-tailed skewed features
wine_df$residual.sugar_log <- log10(wine_df$residual.sugar + 1)
p4 <- ggplot(aes(x = residual.sugar_log), data = wine_df) +
  geom_histogram()

wine_df$chlorides_log <- log10(wine_df$chlorides + 1)
p5 <- ggplot(aes(x = chlorides_log), data = wine_df) +
  geom_histogram()


grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12, ncol = 3)

```

# Bivariate Plots Section

A correlation overview of the bivariate relationships can be seen below. 
Clearly, it appears some attributes have a greater correlation to the quality of 
wine than others do.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Correlation}
# find the correlation between any 2 attributes
cor(wine_df[, 2:13])
```

The correlation matrix gives a preview into what the bivariate scatterplots look 
like between quality and the other features.

## Top Correlated Features (To Quality)

I ran Pearson's correlation tests between the wine quality against the remaining 
wine properties and found 4 features with a correlation coefficient of at least 
0.2  with the wine quality:

- Alcohol, r = 0.476
- Volatile.acidity, r = -0.391
- Sulphates, r = 0.251
- Citric.acid, r = 0.226

## Interesting Relationships Between the Other Features (not the main feature(s) 
## of interest)?

Found strong correlation between:

- Citric.acid vs fixed.acidity , r = 0.672
- Density vs fixed.acidity , r = 0.668
- pH vs fixed.acidity , r = 0.683
- Total.sulfur.dioxide vs free.sulfur.dioxide , r = 0.668

## Bivariate Box Plots - A Bivariate Quality Investigation

Building off of the correlation calculated, box plots were created below to 
observe anythingelse that may have not been captured between the bivariate 
correlation table.

Beginning with the supporting features to the quality feature, that is, the main 
feature of interest, I'm going to explore acidity (fixed, volatile, citric) and 
pH since they could have a big impact on the quality of the wine. It's important
to understand too acidic of wine can give off a strong vinegar taste. Also, wine
with a pH outside the range of 3-4 could be too basic or too acidic and affect 
the taste.

The plots for the attributes of acidity and pH are below,

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Box_Plots_Investigation}
# plot quality vs fixed.acidity
ggplot(aes(x = factor(quality), y = fixed.acidity), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = "mean", geom = "point", color = "red")

# plot quality vs volatile.acidity
ggplot(aes(x = factor(quality), y = volatile.acidity), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs citric.acid
ggplot(aes(x = factor(quality), y = citric.acid), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs pH
ggplot(aes(x = factor(quality), y = pH), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')
```

- Fixed.acidity - As fixed acidity increases or decreases, there's little effect 
to the quality. The mean values of fixed acidity don't fluctuate much as quality 
increases.
- Volatile.acidity - As volatile acidity increases, the quality of wine 
decreases, so there's an inverse relationship between the two.
- Citric.acid - As citric acid increases, there's a slight uptick in wine 
quality.
- pH - As pH decreases (also, becomes more acidic), the quality of wine slightly 
increases as well.

I wanted to explore how the different types of acid affected the pH. Plots 
comparing to the3 types of acid are below.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots_AcidvsPH}
# plot fixed.acidity vs pH
pfixed <- ggplot(aes(x = fixed.acidity, y = pH), data = wine_df) +
  geom_point(alpha = .2) 

# plot volatile.acidity vs pH
pvolatile <- ggplot(aes(x = volatile.acidity, y = pH), data = wine_df) +
  geom_point(alpha = .2) 

# plot citric.acid vs pH
pcitric <- ggplot(aes(x = citric.acid, y = pH), data = wine_df) +
  geom_point(alpha = .2) 

grid.arrange(pfixed, pvolatile, pcitric, ncol = 3)
```

Putting these 3 plots on the same grid made it easier to see how for volatile 
acid the negative correlation that exists between pH and acid doesn't hold true 
here.

Plotted below are the remaining attributes compared to wine quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Box_Plots}
# plot quality vs residual.sugar
ggplot(aes(x = factor(quality), y = residual.sugar), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  scale_y_continuous(lim = c(1, 4)) +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs chlorides
ggplot(aes(x = factor(quality), y = chlorides), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  scale_y_continuous(lim = c(.075, .15)) +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs free.sulfur.dioxide
ggplot(aes(x = factor(quality), y = free.sulfur.dioxide), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  scale_y_continuous(lim = c(0, 100)) +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs total.sulfur.dioxide
ggplot(aes(x = factor(quality), y = total.sulfur.dioxide), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  scale_y_continuous(lim = c(0, 75)) +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs density
ggplot(aes(x = factor(quality), y = density), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality vs sulphates
ggplot(aes(x = factor(quality), y = sulphates), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')
```

- Residual.sugar - Residual sugar content in wine doesn't seem to have any 
affect on the quality as the mean levels stay relatively equal for varying 
quality of wine.
- Chlorides - As chlorides decrease, there's a very minor increase in the 
quality.
- Free.sulfur.dioxide - For free sulfur dioxide, it would appear that absolutely 
low levels on dioxide results in absolutely low levels of quality wine. While, 
absolutely high levels of dioxide results in average quality (~ 4.5 - 6.5).
- Total.sulfur.dioxide - This attribute takes on a similar patter as free sulfur 
dioxide.
- Density - As density decreases, the quality of wine has a slight increase.
- Sulphates - As sulphates increase, the quality of wine increases as well.

Since quality and alcohol held the strongest correlation coefficient of over 0.4 
I decided to perform a deeper investigation of the attribute by also taking a 
look at how the quality_rank attribute performed with varying levels of alcohol.

## Bivariate Plots - Quality vs Alcohol

The box plot of quality vs alcohol demonstrates the strong correlation 
coefficient between the two attributes. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_Quality_Alcohol}
# plot quality vs alcohol
ggplot(aes(x = factor(quality), y = alcohol), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')

# plot quality_rank vs alcohol
ggplot(aes(x = factor(quality_rank), y = alcohol), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red')
```

A puzzling aspect is how there still exists a handful of outliers in the box 
plot. This suggests alcohol isn't the only attribute with a contribution to the 
quality of wine. The bivariate correlation of quality supports this idea. A 
linear model was calculated below to gain insight into just how much other 
attributes and alcohol affect the quality of wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Linear_Models}
# linear model with alcohol as the explanatory variable and quality as the outcome variable
m1 <- lm(as.numeric(quality) ~ alcohol, data = wine_df)
summary(m1)

# linear model with the top 4 correlated variables as the explanatory variable set
m2 <- lm(as.numeric(quality) ~ alcohol + volatile.acidity + sulphates + citric.acid, data = wine_df)
summary(m2)
```

Looking at the R squared value we can see that the alcohol variable contributes 
to 22% of the quality of wine. Adding onto this linear model we were able to 
find that the top 4 variables correlated to quality total 34% of the quality.

Additional box plots are below for the transformed quality attribute and the 
other 3 attributes with strong correlation (correlation coefficient at least 
0.2).

```{r echo=FALSE, message=FALSE, warning=FALSE, Box_Plot_Analysis}
ggplot(wine_df, 
       aes(x=quality_rank, y=volatile.acidity, fill=quality_rank)) + 
       geom_boxplot()

ggplot(wine_df, 
       aes(x=quality_rank, y=sulphates, fill=quality_rank)) + 
       geom_boxplot()

ggplot(wine_df, 
       aes(x=quality_rank, y=citric.acid, fill=quality_rank)) + 
       geom_boxplot()
``` 

Violin for the quality to alcohol content is shown below.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_Violin}
# plot quality vs alcohol
# try a jitter plot or a violin instead since the quality feature is categorical - scatter won't work
ggplot(aes(x = factor(quality), y = alcohol), data = wine_df) +
  geom_violin()
```

This violin plot shows the increasing trend of alcohol content affecting
wine quality.

# Bivariate Analysis

In this analysis, we explore any relationships that may exist between the newly 
transformed quality_rank attribute with the chemical properties. 

## Chemical Properties Distributions Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram}
# alcohol vs quality_rank
ggplot(aes(x = alcohol, fill = quality_rank), data = wine_df) +
  geom_histogram(binwidth = .5, alpha = .5)
```

The plot shows the distribution of alcohol at the different wine quality ranks. 
The data distribution for each quality is not equal (ie there's minimal 
observations for poor tasting wine) so it makes it difficult to visually denote 
disitrbutions from the histogram. A normalization of the plot is seen below.

```{r echo=FALSE, message=FALSE, warning=FALSE, HistogramNormalizedAlcohol}
# pdf of alcohol vs quality_rank
ggplot(aes(x = alcohol, fill = quality_rank), data = wine_df) +
  geom_density(binwidth = .5, alpha = .5)
```

For the remaining 3 attributes of high correlation coefficients - 
volatile.acidity, sulphates, citric.acid - the distributions plots against the 
wine quality at the different ranks are below.

```{r echo=FALSE, message=FALSE, warning=FALSE, HistogramNormalized}
#Volatile.acidity, r = -0.391
#Sulphates, r = 0.251
#Citric.acid, r = 0.226

# volatile.acidity vs quality_rank
ggplot(aes(x = volatile.acidity, fill = quality_rank), data = wine_df) +
  geom_density(binwidth = .5, alpha = .5)

# sulphates vs quality_rank - sulphates is so skewed so try the sqrt since the regular attribute creates a scrunched / cluttered distribution
ggplot(aes(x = sulphates_sqrt, fill = quality_rank), data = wine_df) +
  geom_density(binwidth = .25, alpha = .5)

# citric.acid vs quality_rank
ggplot(aes(x = citric.acid, fill = quality_rank), data = wine_df) +
  geom_density(binwidth = .25, alpha = .5)
```

## Observation Summary of the Bivariate Plots

1. Alcohol has the strongest correlation to the quality of wine, with a 
coefficient = 0.476.
2. Fixed acidity has no effect on quality.
3. Volatile acidity has a negative correlation on quality.
4. pH has a negative effect on quality - meaning, the more acidic a wine is, 
the higher the quality.
5. Residual sugar has no effect on quality.
6. Density has a minor negative effect on quality.
7. Sulphates have a minor positive effect on quality.
8. Alcohol contributes to the quality of wine by a factor of 22%, 
the most out of any of the other variables in the wine data set.

# Multivariate Plots Section

## Effect of Acids on Wine Quality

Based on the bivariate investigation of the role acids play in the quality of 
wine we know that the more acidic a wine is, the higher the quality is. Let's 
investigate how different combinations of acids can affect the quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AcidsOnQuality}
ggplot(aes(x = citric.acid, y = fixed.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Fixed Acidity by Citric Acid and Quality Rank')

ggplot(aes(x = volatile.acidity, y = fixed.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Fixed Acidity by Volatile Acidity and Quality Rank')

ggplot(aes(x = citric.acid, y = volatile.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Volatile Acidity by Citric Acid and Quality Rank')

ggplot(aes(x = citric.acid, y = fixed.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Fixed Acidity by Citric Acid and Quality Rank')
```

- Fixed Acidity by Citric Acid and Quality Rank - There doesn't seem to be any
correlation to the wine quality with these two variables.
- Fixed Acidity by Volatile Acidity and Quality Rank - There doesn't seem to be 
any correlation to the wine quality with these two variables.
- Volatile Acidity by Citric Acid and Quality Rank - Lower amounts of volatile 
acidity
and higher amounts of citric acid lead to better quality wine.
- Fixed Acidity by Citric Acid - There doesn't seem to be any
correlation to the wine quality with these two variables.

## Effect of Alcohol and Other Attributes on Wine Quality

Furthering the investigation into which attributes affect the quality of wine,
I held alcohol as an independent variable to test whether different variables
affected the quality of wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_Fixed.acidity}
# plot of fixed.acidity
ggplot(aes(x = alcohol, y = fixed.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Fixed Acidity by Alcohol and Quality Rank')
```

In this relationship, wine quality increases as fixed acidity content increases 
and alcohol concentration decreases.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_Volatile.acidity}
# plot of volatile.acidity
ggplot(aes(x = alcohol, y = volatile.acidity, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Volatile Acidity by Alcohol and Quality Rank')
```

As the amount of volatile acidity decreases and the amount of alcohol increases, 
the quality of wine seems to increase as well.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_CitricAcid}
# plot of citric.acid
ggplot(aes(x = alcohol, y = citric.acid, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(0, .75)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Citric Acid by Alcohol and Quality Rank')
```

With higher citric acid and lower alcohol concentration, the wine quality 
increases.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_ResidualSugar}
# plot of residual.sugar
ggplot(aes(x = alcohol, y = residual.sugar, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(1,4)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Residual Sugar by Alcohol and Quality Rank')
```

There doesn't seem to be any correlation with here between residual sugar and 
quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_Chlorides}
# plot of chlorides
ggplot(aes(x = alcohol, y = chlorides, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(.03,.175)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Chlorides by Alcohol and Quality Rank')
```

Chlorides behaves in a similar way as residual sugar does with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_FreeSulfurDioxide}
# plot of free.sulfur.dioxide
ggplot(aes(x = alcohol, y = free.sulfur.dioxide, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Free Sulfur Dioxide by Alcohol and Quality Rank')
```

As free sulfur dioxide concentration decreases and alcohol increases, the 
quality of wine decreases.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_pH}
# plot of pH
ggplot(aes(x = alcohol, y = pH, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(3,3.6)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('PH by Alcohol and Quality Rank')
```

As pH decreases and alcohol concentration increases, wine quality improves.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_AlcoholQuality_Sulphates}
ggplot(aes(x = alcohol, y = sulphates, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(.25, 1.2)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Sulphates by Alcohol and Quality Rank')
```

As sulphate concentration increases and alcohol concentration increases, the 
quality of wine increases as well. Interesting to note how excellent wine seems 
to hover around .75 sulphates level even at varying values of alcohol.

# Multivariate Analysis

## Linear Modeling

Using the linear model function I was able to build a model to predict the 
outcome variable, quality, using a handful of explanatory (or predictor) 
variables in the wine dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Linear_Modeling}
# use variables with a correlation coefficient of at least .1
set.seed(1000001)
wine_samp <- wine_df[sample(1:length(wine_df$quality), 500), ]
lm_tot <- lm(as.numeric(quality) ~ fixed.acidity + 
            volatile.acidity +
            chlorides +
            total.sulfur.dioxide +
            sulphates + 
            alcohol, 
          data = wine_samp)

summary(lm_tot)
```

The final linear model for wine quality becomes:

lm(quality) = 3.080 + 0.040*fixed.acidity - 1.039*volatile.acidity - 
2.164*chlorides - 0.002*total.sulfur.dioxide + 0.753*sulphates + 0.244*alcohol

## Analysis of Linear Model

I first ran the linear model using the top 8 correlated variables to quality to
determine the significance of the variables and limit the model to only include 
signifcantly important predictor variables. Using the signifiance legend codes 
as a guide I kept any variable with at least a "*" next to it for the final 
model feature set. The reason I used the signifance codes as a way to interpret 
linear model effectiveness is because the of the low R squared value I kept 
calculating in the various linear models. It would seem that the vast majority 
of wine observations belonging to the average quality rank could come into play 
here as a more complete data set - one with a more level-loaded wine quality - 
could have aided in developing a stronger model.

# Final Plots and Summary

In determining the most significant plots in this investigation on the red wine 
dataset I approached it from a multi-tiered level since at each section in this 
investigation, I gleaned insight from previous sections - univariate to 
bivariate analysis, and bivariate to multivariate analysis.

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# histogram for quality_rank
ggplot(aes(x = quality_rank), data = wine_df) +
  geom_bar(width = .5, 
                 color = I('black'), 
                 fill = I('#099DD9')) +
  ggtitle('Distribution of Wine Quality') +
  xlab('Wine Quality') + ylab('Amount of Observations')
```

### Description One

I chose plot one because it shows the obvious unbalanced distribution of the 
dataset in that the bulk of the data falls into the average wine quality bucket. 

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# plot quality vs alcohol
ggplot(aes(x = factor(quality), y = alcohol), data = wine_df) +
  geom_boxplot(alpha = .5, color = 'blue') +
  geom_jitter(alpha = .2) +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  ggtitle('Wine Quality Rank by Alcohol') +
  xlab('Wine Quality Rank') + ylab('Alcohol [%]')
```

### Description Two

This next plot demonstrated that the amount of alcohol concentration in a 
particular wine can be a good indicator into the quality of that wine. It's easy 
to see that as the percentage of alcohol climbs, there exists a trend for wine 
quality to climb as well.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = sulphates, colour = quality_rank), data = wine_df) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_y_continuous(lim = c(.25, 1.2)) +
  geom_smooth(method = 'lm', se = FALSE, size = 1.25) +
  scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality')) +
  ggtitle('Sulphates by Alcohol and Quality Rank') +
  xlab('Alcohol [%]') + ylab('Sulphates')
```

### Description Three

This last plot is of a multivariation relationship demonstrating how better wine 
quality can be achieved through increased levels of alcohol and sulphates. 
Previously in the investigation, we knew that alcohol contributed to 22% of the 
quality of wine. This plot showed that, at least part of the remaining 78%, 
included sulphates as a contributing determinant to predicting the quality of 
wine.

------

# Reflection

This red wine dataset contains 1599 observations with 11 variables on various 
chemical properties. With the main feature of interest being the quality of 
wine, this investigation's objective was to explore the relationship between the 
rest of the features and the wine quality. 

Initial bivariate exploration revealed that only a handful of features hold a 
reasonable correlation to the wine quality. For the purposes of this 
investigation, the top 4 most highest correlation coefficients to quality - 
alcohol, volatile acidity, sulphates, and citric acid - were used in much of 
the data visualizations.

It was surprising to me how some of the other features aren't strongly 
correlated with the wine quality. I would assume pH, for example, would play a 
big role into determining the quality of wine.

Another surprise / struggle appeared when I realized the uneven distribution of 
observations across the 3 levels of wine quality ranking - poor, average, 
excellent. We saw how in the linear model for predicting wine quality, the 
traditional approach for using R squared values as a gauge for model 
effectiveness did not work with this particular dataset (since all the values of 
models were well below acceptance ranges). A data collection effort to include 
more wines from poor and excellent rankings would go a long way in improving the 
capability of models for predicting the quality of wine.

In terms of the future work to be done for improving the dataset, including 
different components of red wine into the data could help lead to more powerful 
linear models as well. For example, I'm curious to see how age would affect the 
quality of wines. Is older-aged wine really worth the bump in retail tag? Or 
does the price of wine help determine the quality? These questions could help 
spark an effort to form a dataset with the potential to gain insight and strong 
conclusions from.


















